---
title: "MiniZinc Interface for R"
author: "Akshit Achara, Lars Kotthoff" 
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MiniZinc Interface for R}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


## Introduction

MiniZinc is a free and open-source constraint modeling language. Constraint satisfaction and discrete optimization problems can be formulated in a high-level modeling language. Models are compiled into an intermediate representation that is understood by a wide range of solvers. MiniZinc itself provides several solvers, for instance GeCode. The existing packages in R are not powerful enough to solve even mid-sized problems in combinatorial optimization.

There are implementations of an Interface to MiniZinc in Python like MiniZinc Python and pymzn and JMiniZinc for Java but such an interface does not exist for R.

This package provides an implementation of a very simple and easy to use interface for R that will help R users to solve optimization problems that can't be solved with R currently.

It's important to understand R6 classes before getting into the details. If you are not comfortable with R6, please go through this [tutorial](https://adv-r.hadley.nz/r6.html).

It would be nice to go through this [manual](https://www.minizinc.org/tutorial/minizinc-tute.pdf) to understand more about MiniZinc. This is mainly for those who are interested in contributing to the package.

Please make sure to install MiniZinc on your operating system and also ensure that minizinc is accessible from command line by typing 'minizinc' on the command line without the single quotes. The installation instructions can be found [here](https://www.minizinc.org/doc-2.4.3/en/installation.html).

## Getting Started

```{r, results='hide', echo=FALSE}
library(rminizinc)
```

### Parse a MiniZinc Model

A parser function `mzn_parse` has been implemented which can be used to detect possible syntax errors and get the smallest of details before the MiniZinc model is evaluated. The function returns the initialize `Model` R6 object.


Now, let's solve a job shop model:

```{r}
# mzn file path
mzn_path = paste0(dirname(getwd()), "/inst/extdata/mzn_examples/jobshop/jobshop_0.mzn")

# parse the model
parseObj=rminizinc:::mzn_parse(mzn_path = mzn_path)
```

Look at the contents of parseObj for more understanding of the model.

### Find the Missing Parameters

The missing parameters can be obtained using `get_missing_pars()`

```{r}
missingPars = get_missing_pars(model = parseObj)
print(missingPars)
```

### Set Values of Missing Parameters 

```{r}
pVals = list(Int$new(3), Int$new(4),
             Array$new(exprVec = intExpressions(c(3, 3, 4, 4, 4, 3, 2, 2, 3, 3, 3, 4)),
               dimranges = list(IntSetVal$new(1,3), IntSetVal$new(1,4))), 
             Array$new(exprVec = intExpressions(c(1, 2, 3, 4, 1, 3, 2, 4, 4, 2, 1, 3)),
               dimranges = list(IntSetVal$new(1,3), IntSetVal$new(1,4))))
names(pVals) = missingPars
model = set_params(model = parseObj, modData = pVals)
cat(model$mzn_string())
```

### Solve the Model

The function `mzn_eval()` is used to evaluate a MiniZinc model and returns the solution string and a list of solutions if they were parsed without any error by the function `sol_parse()` otherwise it returns the solution string and an appropriate error. The parsed solutions are a named list where elements are of type `OBJ$SOLUTIONS$SOLUTION<n>$<VARIABLE_NAME>`. The optimal solution if found can be accessed using `OBJ$SOLUTIONS$OPTIMAL_SOLUTION` and the best solution can be accessed using `OBJ$SOLUTIONS$BEST_SOLUTION`. More details about the functions can be obtained using `?mzn_eval` and `?sol_parse`.

The solver that should be used to solve the model needs to be specified by the user (default is Gecode)
and the lib_path (path of the MiniZinc executable) should also be provided. If the user wishes to provide a data file, it's path can be provided to the argument `dznpath`.  A time limit (in ms) can also be provided to the argument `time_limit`. (default is 300000 ms)

A sample job shop problem has been solved below:

```{r}
# R List object containing the solutions
solObj = rminizinc:::mzn_eval(model, solver = "org.gecode.gecode",
                   lib_path = "/snap/minizinc/current/share/minizinc")
# get all the solutions
print(solObj$SOLUTIONS)
```

Let's solve another problem.

```{r}
# file path
mzn_path = paste0(dirname(getwd()), "/inst/extdata/mzn_examples/knapsack/knapsack_0.mzn")

# get missing parameter values
missingVals=rminizinc:::get_missing_pars( model = mzn_parse(mzn_path = mzn_path))
print(missingVals)

# list of the data
pVals = list(Int$new(3), Int$new(9), Array$new(intExpressions(c(15,10,7)))
             , Array$new(intExpressions(c(4,3,2))))
names(pVals) = missingVals

# set the missing parameters
model = rminizinc:::set_params(modData = pVals, 
                                   mzn_parse(mzn_path = mzn_path))

# R List object containing the solutions
solObj = rminizinc:::mzn_eval(model, solver = "org.gecode.gecode",
                     lib_path = "/snap/minizinc/current/share/minizinc")
# get all the solutions
print(solObj$SOLUTIONS)

```

Some examples of how to use these functions to solve optimization problems can be found in `mzn_examples` which are taken from [minizinc-examples](https://github.com/MiniZinc/minizinc-examples).

NOTE: Please don't include output formatting in the mzn files or the solutions might not be parsed properly.

## Variables

There are two types of variables in MiniZinc namely, decision variables and parameters. 

The data types of variables can be single types i.e integers (int), floating point numbers (float), Booleans (bool) and strings (string) and collections i.e sets, enums and arrays (upto 6 dimensional arrays).

Parameter is used to specify a parameter in a given problem and they are assigned a fixed value or expression.

Decision variables are the unknowns that Minizinc model is finding solutions for. We do not need to give them a value, but instead we give them a domain of possible values. Sometimes expressions involving other variables and parameters are also assigned to decision variables. Decision variables need to satisfy a set of constraints which form the core of the problem. 

To create a variable declaration one needs to understand the elements of R6 classes that will be used to create the variables.

Easy to use declaration functions have been created for the users to declare variables and parameters of different data types. Examples of how to declare variables is shown below.

```{r}
# create the variable and parameter declarations
item1 = VarDeclItem$new(decl = IntDecl(name = "n", kind = "par"))

par2_val = BinOp$new(lhs = Int$new(1), binop = "..", rhs = item1$id())
item2 = VarDeclItem$new(decl = IntSetDecl(name = "OBJ", kind = "par", value = par2_val))

item3 = VarDeclItem$new(decl = IntDecl(name = "capacity", kind = "par"))

item4 = VarDeclItem$new(decl = IntArrDecl(name = "profit", kind = "par", ndim = 1, 
                                          ind = list(item2$id())))

item5 = VarDeclItem$new(decl = IntArrDecl(name = "size", kind = "par", ndim = 1, ind =                                                            list(item2$id())))

item6 = VarDeclItem$new(decl = IntArrDecl(name = "x", kind = "var", ndim = 1, ind = list(item2$id())))
```

## Constraints

Constraints are defined on the decision variables to restrict the range of values that they can take. They can also be thought of as the rules of a problem.

Constraints can be created using different R6 sub classes of the super class Expression. In this example Generator, Binop, Comprehension and Call classes have been used. These classes take in the elements required to create an expression that will be used as a constraint. More information can be found using `?<class Name>`

Create constraints:

```{r}

# declare parameter for iterator
parIter = IntDecl(name = "i", kind = "par")


gen_forall = Generator$new(IN = item2$id(), decls = list(parIter))
bop1 = BinOp$new(lhs = ArrayAccess$new(v = item6$id(),  args= list(gen_forall$getDecl(1))),
                                                             binop = ">=", rhs = Int$new(0))

Comp1 = Comprehension$new(generators = list(gen_forall), body = bop1, set = FALSE)
cl1 = Call$new(fnName = "forall", args = list(Comp1))
item7 = ConstraintItem$new(e = cl1)

gen_sum = Generator$new(IN = item2$id(), decls = list(parIter))

bop2 = BinOp$new(lhs = ArrayAccess$new(v = item5$id(), args = list(gen_sum$getDecl(1))),                  
                 binop = "*",  rhs = ArrayAccess$new(v = item6$id() , 
                 args = list(gen_sum$getDecl(1))))

Comp2 = Comprehension$new(generators = list(gen_sum), body = bop2, set = FALSE)
cl2 = Call$new(fnName = "sum", args = list(Comp2))
bop3 = BinOp$new(lhs = cl2, binop = "<=", rhs = item3$id())
item8 = ConstraintItem$new(e = bop3)
```


## Solve Type

The constraint programming problem can be of three types, namely: Satisfaction , Minimization and Maximization.
Satisfaction problems produce all the solutions that satisfy the constraints whereas minimization and maximization problems produce the solution which minimizes and maximizes the given expression.

An example is shown below:

```{r}

bop4 = BinOp$new(lhs = ArrayAccess$new(v = item4$id(), args = list(gen_sum$getDecl(1))),
                      binop = "*", rhs = ArrayAccess$new(v = item6$id(), 
                      args = list(gen_sum$getDecl(1))))

Comp3 = Comprehension$new(generators = list(gen_sum), body = bop4, set = FALSE)

cl3 = Call$new(fnName = "sum", args = list(Comp3))

item9 = SolveItem$new(solve_type = "maximize", e = cl3)
```

## Create the Model

Combine all the items to create a MiniZinc model.

```{r}
items  = c(item1, item2, item3, item4, item5, item6, item7, item8, item9)
mod = Model$new(items = items)
modString = mod$mzn_string()
cat(modString)
```

## Create Items Using Strings 

The strings containing MiniZinc syntax of items can be directly supplied to the constructors to initialize the objects. If strings are supplied, no other argument should be supplied to any of the Item classes except for `AssignItem` where you need to provided the associated variable declaration for the assignment.

### Varaible Item

```{r, results = 'hold'}
declItem = VarDeclItem$new(mzn_str = "set of int: WORKSHEET = 0..worksheets-1;")
sprintf("Is this a parameter? %s", declItem$getDecl()$isPar())
sprintf("Is this a set? %s", declItem$getDecl()$ti()$type()$isSet())
sprintf("Base type of set: %s", declItem$getDecl()$ti()$type()$bt())
sprintf("Name: %s", declItem$id()$getId())
sprintf("Value: %s", declItem$getDecl()$getValue()$c_str())
```

### Constraint Item

```{r, results = 'hold'}
CstrItem = ConstraintItem$new(mzn_str = "constraint forall (i in PREC)
                  (let { WORKSHEET: w1 = preceeds[i];
		                     WORKSHEET: w2 = succeeds[i]; } in
                   g[w1] * e[w1] <= d[w2] + days * (1 - g[w2]));")
sprintf("Expression involved: %s", CstrItem$getExp()$c_str())
sprintf("Call function name: %s", CstrItem$getExp()$getName())
sprintf("Number of Arguments: %s", CstrItem$getExp()$nargs())
sprintf("Class of Argument: %s",  class(CstrItem$getExp()$getArg(1))[1])
sprintf("Number of Generators: %s", CstrItem$getExp()$nargs())
sprintf("Generator: %s", CstrItem$getExp()$getArg(1)$getGen_i(1)$c_str())
sprintf("Comprehension body: %s", CstrItem$getExp()$getArg(1)$getBody()$c_str())
```

### Solve Item

```{r, results = 'hold'}
SlvItem = SolveItem$new(mzn_str = "solve 
    :: int_search(
        [ if j = 1 then g[import_first[i]] else -d[import_first[i]] endif  | i in 1..worksheets, j in 1..2], 
        input_order, indomain_max, complete)
    maximize objective;")
sprintf("Objective: %s", SlvItem$getSt())
cat(sprintf("Annotation: %s", SlvItem$getAnn()$c_str()))
```

### Function Item

```{r, results = 'hold'}
fnItem = FunctionItem$new(mzn_str = "predicate nonoverlap(var int:s1, var int:d1,
                     var int:s2, var int:d2)=
          s1 + d1 <= s2 \\/ s2 + d2 <= s1;")
sprintf("Function name: %s", fnItem$name())
sprintf("No of function declarations: %s", length(fnItem$decls()))
sprintf("Function expression: %s", fnItem$body()$c_str())
```

### Include Item

```{r}
iItem = IncludeItem$new(mzn_str = "include \"cumulative.mzn\" ;")
sprintf("Included mzn name: %s", iItem$getmznName())
```

## Delete Items

All the item classes have a `delete()` function which is used to delete the item. An example to demonstrate the same is shown below:

```{r}
# delete the item 1
item1$delete()
# check that item1 has been deleted
ls(pattern = "item[^a-z]")
# item1 reference is deleted but items vector should be updated
items = c(item2, item3, item4, item5, item6, item7, item8, item9)
mod = Model$new(items)
cat(mod$mzn_string())
```

## Modify Variable Domain

```{r, results = 'hold'}
vd = VarDomainDecl(name = "n", dom = Set$new(IntSetVal$new(imin = 1, imax = 2)))
sprintf("The current declaration is: %s", vd$c_str())
vd$setDomain(Set$new(IntSetVal$new(imin = 3, imax = 5)))
sprintf("The modified declaration is: %s", vd$c_str())
```

## Modify Constraints

There are various getter and setter functions for the expression classes that can be used to modify existing constraints. For example:

```{r, results = 'hold'}
vItem = VarDeclItem$new(mzn_str = "set of int: a = {1, 2, 3, 4};") 
cItem = ConstraintItem$new(mzn_str = "constraint sum(a) < 10;")
sprintf("The current constraint is: %s", cItem$c_str())
cItem$setExp(BinOp$new(lhs = Call$new(fnName = "max", args = list(vItem$getDecl()$id())),
                       binop = "<", rhs = Int$new(10)))
sprintf("The modified constraint is: %s", cItem$c_str())
```