---
title: "MiniZinc Interface for R"
author: Akshit Achara 
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R functionalities to use MiniZinc}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


## Introduction

MiniZinc is a free and open-source constraint modeling language. Constraint satisfaction and discrete optimization problems can be formulated in a high-level modeling language. Models are compiled into an intermediate representation that is understood by a wide range of solvers. MiniZinc itself provides several solvers, for instance GeCode. The existing packages in R are not powerful enough to solve even mid-sized problems in combinatorial optimization

There are implementations of an Interface to MiniZinc in Python like MiniZinc Python and pymzn and JMiniZinc for Java but such an interface does not exist for R.

This package provides an implementation of a very simple and easy to use interface for R that will help R users to solve optimization problems that can't be solved with R currently.

It's important to understand R6 classes before getting into the details. If you are not comfortable with R6, please go through this [tutorial](https://adv-r.hadley.nz/r6.html).

It would be nice to go through this [manual](https://www.minizinc.org/tutorial/minizinc-tute.pdf) to understand more about MiniZinc. This is mainly for those who are interested in contributing to the package.

Please make sure to install MiniZinc on your operating system and also ensure that minizinc is accessible from command line by typing 'minizinc' on the command line without the single quotes. The installation instructions can be found at [instructions](https://www.minizinc.org/doc-2.4.3/en/installation.html)


## Variables

There are two types of variables in MiniZinc -- decision variables and parameters. 

The data types of variables can be single types i.e integers (int), floating point numbers (float), Booleans (bool) and strings (string) and collections i.e sets, enums and arrays (upto 6 dimensional arrays).

Parameter is used to specify a parameter in a given problem and they are assigned a fixed value or expression.

Decision variables are the unknowns that Minizinc model is finding solutions for. We do not need to give them a value, but instead we give them a domain of possible values. Sometimes expressions involving other variables and parameters are also assigned to decision variables. Decision variables need to satisfy a set of constraints which form the core of the problem. 

To create a variable declaration one needs to understand the elements of R6 classes that will be used to create the variables.

The class `VarDecl` can be represented as:

```{r VarDecl UML, echo=FALSE, out.width = '30%'}
knitr::include_graphics(paste0(getwd(),"/workflows/VarDecl.png"))
```

The type information of the declaration is stored in an object of class Type which holds information like base type, number of dimensions, etc. The type instantiation is stored in an object of TypeInst class which contains the type information as well as the index or domain in case the variable is an array or decision variable with domain respectively. The value assigned to the variable declaration is stored in an object of sub classes of super class Expression that can be Int, Float, Array, or other Expression sub classes that result in a value of corresponding type.

The UML diagram of Expression class can be seen below:
```{r Expression UML, echo=FALSE, out.width = '80%'}
knitr::include_graphics(paste0(getwd(),"/workflows/ExpressionUml.png"))
```

Here are examples of how we can create parameters and decision variables.

```{r}
library(rminizinc)

# create the variable and parameter declarations
par_ti = TypeInst$new(Type$new(base_type = "INT", kind = "parameter"))
par1 = VarDecl$new(id = "n", par_ti, type_inst = par_ti)
# create the Item 1
item1 = VarDeclItem$new(decl = par1)

sv = SetVal$new(val = c(l = 1, u=par1))
par2_dt = Set$new(setVal = sv)
par2_ti = TypeInst$new(Type$new(base_type = "INT", kind = "parameter", dim = 1, set_type = TRUE))
par2 = VarDecl$new(type_inst = par2_ti, expression = par2_dt, id = "OBJ")
# create the item 2
item2 = VarDeclItem$new(decl = par2)

par3_ti = TypeInst$new(Type$new(base_type = "INT", kind = "parameter"))
par3 = VarDecl$new(type_inst = par3_ti, id = "capacity")
# create the item 3
item3 = VarDeclItem$new(decl = par3)

par4_ti = TypeInst$new(Type$new(base_type = "INT", kind = "parameter", dim = 1),
                   indexExprVec = par2$id())
par4 = VarDecl$new(type_inst = par4_ti, id = "profit")
# create the item 4
item4 = VarDeclItem$new(decl = par4)

par5_ti = TypeInst$new(Type$new(base_type = "INT", kind = "parameter", dim = 1),
                   indexExprVec = par2$id())
par5 = VarDecl$new(type = par5_ti, id = "size")
# create the item 5
item5 = VarDeclItem$new(decl = par5)

par6_ti = TypeInst$new(Type$new(base_type = "INT", kind = "decision", dim = 1),
                   indexExprVec = par2$id())
par6 = VarDecl$new(type = par6_ti, id = "x")

#create the item 6
item6 = VarDeclItem$new(decl = par6)
```

## Constraints

Constraints are defined on the decision variables to restrict the range of values that they can take. They can also be thought of as the rules of a problem.

Constraints can be created using different R6 sub classes of the super class Expression. In this example Generator, Binop, Comprehension and Call classes have been used. These classes take in the elements required to create an expression that will be used as a constraint. More information can be found using `?<class Name>`

Create constraints:

```{r}
# generator
gen_forall = Generator$new(IN = par2$id(), iterator = "i")
# binary operator expression
bop1 = Binop$new(lhs_expression = ArrayAccess$new(id = par6$id(), index = gen_forall$iter_id()), binop = ">=", 
                                        rhs_expression =    Int$new(value = IntVal$new(0)))
# comprehension
Comp1 = Comprehension$new(generators = list(gen_forall), expression = bop1)
# forall function call
cl1 = Call$new(fn_id = "forall", lExp = list(Comp1))
# item7 
item7 = ConstraintItem$new(expression = cl1)

gen_sum = Generator$new(IN = par2$id(), iterator = "i")
bop2 = Binop$new(lhs_expression = ArrayAccess$new(id = par5$id(), index = gen_sum$iter_id()), binop = "*", 
                                      rhs_expression = ArrayAccess$new(id = par6$id() ,index = gen_sum$iter_id()))
Comp2 = Comprehension$new(generators = list(gen_sum), expression = bop2)
cl2 = Call$new(fn_id = "sum", lExp = list(Comp2))
bop3 = Binop$new(lhs_expression = cl2, binop = "<=", rhs_expression = par3$id())
# item8
item8 = ConstraintItem$new(expression = bop3)
```


## Solve Type

The constraint programming problem can be of three types, namely: Satisfaction , Minimization and Maximization.
Satisfaction problems produce all the solutions that satisfy the constraints whereas minimization and maximization problems produce the solution which minimizes and maximizes the given arithmetic expression.

An example is shown below:

```{r}

gen_sum = Generator$new(IN = par2$id(), iterator = "i")

bop4 = Binop$new(lhs_expression = ArrayAccess$new(id = par4$id(),index = gen_sum$iter_id()), 
                      binop = "*", rhs_expression = ArrayAccess$new(id = par6$id(), index = gen_sum$iter_id()))

Comp3 = Comprehension$new(generators = list(gen_sum), expression = bop4)

cl3 = Call$new(fn_id = "sum", lExp = list(Comp3))

# item9
item9 = SolveItem$new(solve_type = "maximize", expression = cl3)
```

## Create the model

Currently only three types of items can be created in a MiniZinc model namely: Variable Declarations, Constraints and Solve Items. The model is created by combining these items.

The UML diagram is shown below:
```{r Model UML, echo=FALSE, out.width = '70%'}
knitr::include_graphics(paste0(getwd(),"/workflows/ItemUml.png"))
```

Combine all the items to create a MiniZinc model
```{r}
items  = c(item1, item2, item3, item4, item5, item6, item7, item8, item9)
mod = Model$new(items = items)
modString = mod$mzn_string()
```

## Solve the model

Solve the model using the Rcpp functions that are using the MiniZinc.

```{r}
dzn_path = paste0(dirname(getwd()), "/mzn_examples/knapsack/knapsack_0.dzn")
sol = rminizinc:::mzn_eval(modelString = modString, solver = "org.gecode.gecode",
                      libpath = "/snap/minizinc/current/share/minizinc", dznpath = dzn_path)
```


A Minizinc implementation of the same problem is given below:

```{minizinc}

int: n; % number of objects
set of int: OBJ = 1..n;
int: capacity;
array[OBJ] of int: profit;
array[OBJ] of int: size;

array[OBJ] of var int: x; % how many of each object

constraint forall(i in OBJ)(x[i] >= 0);
constraint sum(i in OBJ)(size[i] * x[i]) <= capacity;
solve maximize sum(i in OBJ)(profit[i] * x[i]);

```

Currently only the knapsack problem specified above can be created using the current implementation. More problems will be supported soon.

Now, that the basic structure of the problem is clear, we can create functions for solving different constraint programming problems.

The workflow of the approach used above is given below:
```{r Workflow 1, echo=FALSE, out.width = '80%'}
knitr::include_graphics(paste0(getwd(),"/workflows/write_model.png"))
```

This approach is helpful for R users who have little understanding of MiniZinc.

The next approach is to use the MiniZinc C++ API. This involves giving the MiniZinc model to the Rcpp function, evaluate the results and return them back to R. The workflow can be seen below:
```{r Workflow 2, echo=FALSE, out.width = '80%'}
knitr::include_graphics(paste0(getwd(),"/workflows/API.png"))
```

## Parse Model, Set Missing Parameters and Solve 

A parser function `mzn_parse` has also been implemented which can be used to detect possible syntax errors and get the smallest of details before the MiniZinc model is evaluated. This function can either take the path of a model file path or it can take a string model and returns a list with different elements like: `MODEL_STRING` which is the string representation of the model and `VARIABLES` which are the variable declaration details, `CONSTRAINTS` which are the constraint details of a model, `SOLVE_TYPE` which contains the information about the solve type of the problem, `FUNCTION_ITEMS` contains the function information, `ASSIGNMENTS` contains the assignment information, `INCLUDES` contains information about the included mzn files and `OUTPUT_ITEM` contains the information about the output formatting. Each component (Variable Declarations, Constraints, Solve Items, Assignment Items, Include Items, Functions and Assignments) in a MiniZinc model is considered as an item. The `ITEM_NO` for each element is also returned.

The Information about the naming of the elements returned is described in the following figures:

### Model Items

*Name*: `VARIABLES`    
*Components*: `DECL<n>` is the list of declaration with components `DETAILS` and `ITEM_NO`  
*Description*: `DETAILS` are the details of a variable declaration namely: `NAME`, `TYPE`, `KIND`, `DOMAIN` and `VALUE` and `ITEM_NO` is the item number of the variable declaration in the model.  

*Name*: `CONSTRAINTS`  
*Components*: `CONSTRAINT<n>` is the list of constraints with components: `DETAILS` and `ITEM_NO`.    
*Description*: `DETAILS` are the details about the expressions.

*Name*: `FUNCTION_ITEMS`  
*Components*: `FUNCTION<n>` is the list of functions with components: `DETAILS` and `ITEM_NO`.    
*Description*: `DETAILS` are the details that can `EXPRESSION`, `ANNOTATION` or `PARAMETER_DECLARATIONS`.    

*Name*: `SOLVE_TYPE`  
*Components*: `DETAILS`, `OBJECTIVE` and `ITEM_NO`  
*Description*: `DETAILS` are the details that can `EXPRESSION` or `ANNOTATION`, `OBJECTIVE` can be "satisfy", "minimize" or "maximize".  

*Name*: `ASSIGNMENTS`  
*Components*: `ASSIGNMENT<n>` have the sub components: `DETAILS` and `ITEM_NO`.  
*Description*: `DETAILS` contain the value expression details.

*Name*: `INCLUDES`  
*Components*: `INCLUDED_MZN` and `ITEM_NO`
*Description*: `INCLUDED_MZN` is the name of the included mzn file.

*Name*: `OUTPUT_ITEM`  
*Components*: `DETAILS` and `ITEM_NO`.  
*Description*: `DETAILS` are the details about the output formatting expression.


### Basic Expressions

These are the expressions that lie at the lowest level when parsing any expression.

```{r BASIC EXPRESSIONS, echo=FALSE, out.width = '80%'}
knitr::include_graphics(paste0(getwd(),"/workflows/Basic_Types.png"))
```

### Parent_Expressions

These are the expressions that are above the basic types in hierarchy and can have sub expressions that can be either basic expressions or parent expressions.

```{r PARENT EXPRESSIONS, echo=FALSE, out.width = '80%'}
knitr::include_graphics(paste0(getwd(),"/workflows/Parent_Types.png"))
```

#### Properties of Parent Expressions

Each parent has different properties that define it's structure. The properties of each of the parent are described below:

*Name*: `COMPREHENSION`  
*Components*: `GENERATORS` and `EXPRESSION`  
*Description*: `GENERATORS` is a list of expressions and each generator contains `DECLARATIONS` (list of variable declarations), `IN_EXPRESSION` and `WHERE_EXPRESSION` (Expressions). and `EXPRESSION` is the expression in the comprehension.

*Name*: `BINARY_OPERATION`   
*Components*: `LHS`, `BINARY_OPERATOR` and `RHS`  
*Description*: `LHS` and `RHS` are both expressions (basic or parent) and `BINARY_OPERATOR` is a binary operator in MiniZinc.


*Name*: `UNARY_OPERATION`   
*Components*: `UNARY_OPERATOR` and `ARGUMENTS`   
*Description*: `UNARY_OPERATOR` are the unary operators in MiniZinc and `ARGUMENTS` contain a list of expressions.

*Name*: `VARIABLE_DECLARATION`  
*Components*: `NAME`, `KIND`, `TYPE`, `DOMAIN` and `VALUE`  
*Description*: `NAME`, `KIND` and `TYPE` are strings representing the name, kind and type of the variable and `DOMAIN` and `VALUE` are expressions representing domain and value of the variable declaration.

*Name*: `ARRAY_ACCESS`   
*Components*: `NAME` and `ARGUMENTS`   
*Description*: `NAME` is the name of the variable and `ARGUMENTS` is the list of expressions that represent the indices. 

*Name*: `IF_THEN_ELSE`  
*Components*: `IF`, `THEN` and `ELSE`   
*Description*: `IF` is the list of if expressions, `THEN` is the list of then expressions and `ELSE` is the else expression. 

*Name*: `LET`  
*Components*: `LET_EXPRESSION` and `IN_EXPRESSION`  
*Description*: `LET_EXPRESSION` and `IN_EXPRESSION` are lists of expressions. 

*Name*: `CALL`  
*Components*: `NAME` and `ARGUMENTS`  
*Description*: `NAME` is the name of the function call and `ARGUMENTS` is the list of expressions to that function call.

*Name*: `SET`  
*Components*: `MIN`, `MAX` (set ranges) or `ELEMENT<n>` where n is the serial number of the element.  
*Description*: `MIN` , `MAX` and `ELEMENT<n>` are all expression types (mostly basic expressions).

*Name*: `ARRAY`  
*Components*: `ELEMENT<n>` where n is the serial number of the element.  
*Description*: `ELEMENT<n>` is an expression.

The function `set_params` can be used to set the missing parameters of the model. Any number of missing parameters can be set. The missing parameters can be obtained using `getMissingPars()`. 


```{r}
# mzn file path
mzn_path = paste0(dirname(getwd()), "/mzn_examples/knapsack/knapsack_0.mzn")

# parse the model
parseObj=rminizinc:::mzn_parse(mznpath = mzn_path)

# get the modelString
modString = parseObj$MODEL_STRING

# dzn file path
dzn_path = paste0(dirname(getwd()), "/mzn_examples/knapsack/knapsack_0.dzn")

# R List object containing the solutions
solObj = rminizinc:::mzn_eval(modelString = modString, solver = "org.gecode.gecode",
                     libpath = "/snap/minizinc/current/share/minizinc", dznpath = dzn_path)
# get all the solutions
print(solObj$SOLUTIONS)
```

The data can also be provided as a list of values of parameters where the names of the list must be the names(whole or part) of the parameters whose values are missing. The names of the parameters whose values are missing can be obtained using the function `getMissingPars`. Alternatively, if the user wishes to provide a data file, it can be provided to the function `mzn_eval()`. 

```{r}
# file path
mzn_path = paste0(dirname(getwd()), "/mzn_examples/knapsack/knapsack_0.mzn")

# get missing parameter values
missingVals=rminizinc:::getMissingPars( mznpath = mzn_path)
print(missingVals)

# list of the data
pVals = list(3,9,c(15,10,7),c(4,3,2))
names(pVals) = missingVals

# set the missing parameters
modString = rminizinc:::set_params(modData = pVals,mznpath = mzn_path, modify_mzn = FALSE)

# R List object containing the solutions
solObj = rminizinc:::mzn_eval(modelString = modString, solver = "org.gecode.gecode",
                     libpath = "/snap/minizinc/current/share/minizinc")
# get all the solutions
print(solObj$SOLUTIONS)

```

Some examples of how to use these functions to solve optimization problems can be found in `mzn_examples` which are taken from [minizinc-examples](https://github.com/MiniZinc/minizinc-examples).

NOTE: Please don't include output formatting in the mzn files or the solutions might not be parsed properly.

The next step is to allow the users to modify a MiniZinc model.

## Delete Items

The items/components in a model can be deleted using the function `deleteItem()`. The details about the function and its arguments can be seen using `?deleteItem`. The `ITEM_NO` returned by `mzn_parse()` for each component can be used to delete the required component.The update string representation after deleting the item is returned. The demonstration of how to use this function is shown below:

```{r}
# file path
mzn_path = paste0(dirname(getwd()), "/mzn_examples/knapsack/knapsack_0.mzn")

# parse the model
parseObj=rminizinc:::mzn_parse(mznpath = mzn_path)

print("The original model:")
cat(parseObj$modelString)

# delete the first variable declaration
item_number  = parseObj$VARIABLES$DECL1$ITEM_NO

# delete the item but don't update the mzn
print("The updated model string is:")
cat(deleteItem(itemNo = item_number, mznpath = mzn_path, updateMZN = FALSE))
```


## Modify Domain

The domain of a variable can be modified using the function `modifyDomain`. The item number of the variable whose domain is to be changed needs to be supplied along with the values to be modified. 

The modifications given below are just examples and the modified models aren't meant to be solved.

### minId..maxId 

```{r}

# model string
modString = "int: n; array [1..n] of var 0..1: testVar;"

print("The updated model string is:")
cat(modifyDomainId(ItemNo = 1, maxIdItem = 0, modelString = modString))

```

### min(Integer or Float)..max(Integer or Float)

```{r}
# model string
modString = "int: n; array [1..n] of var 0..1: testVar;"

print("The updated model string is:")
cat(modifyDomainSetVal(ItemNo = 1, imax = 2, imin = 0, modelString = modString))

```


### min(Function Call )..max(Function Call)

```{r}
# model string
modString = "int: n; array [1..n] of var 0..1: testVar;"

print("The updated model string is:")
cat(modifyDomainFnCall(ItemNo = 1, maxIdItem = 0 ,maxFnName = "max", modelString = modString))

```


### Using arithmetic operators and constants


```{r}
# model string
modString = "set of int: n; array [1..n] of var 0..n: testVar;"

print("The updated model string is:")
cat(modifyDomainAO(ItemNo = 1, minVal = 1, maxVal = 1 ,OPmax ='MULT', OPmin = "MULT", modelString = modString))

```